---
title: "Parameter Estimation of the Geometric Model of Visual Meteor Magnitudes"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        toc: true
        fig_width: 6
        fig_height: 4
vignette: >
  %\VignetteIndexEntry{Parameter Estimation of the Geometric Model of Visual Meteor Magnitudes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(vismeteor)
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

## Introduction

The geometric model of visual meteor magnitudes is a commonly used statistical approach
to describe the magnitude distribution of a meteor shower. The observable magnitude distribution
of meteors is then
$$
P[M = m] \sim
\begin{cases}
  f(m_{\mathrm{lim}} - m)\, r^m, & \text{if } m_{\mathrm{lim}} - m > -0.5,\\[5pt]
  0 & \text{otherwise,}
\end{cases}
$$

where $m_{\mathrm{lim}}$ denotes the limiting (non-integer) magnitude of the observation,
and $m$ the integer meteor magnitude.
The function $f(\cdot)$ denotes the perception probability function.

The estimation of the population index r, briefly called the r-value,
is a common task in the evaluation of meteor magnitudes.
Here we demonstrate several methods to estimate this parameter.

First, we obtain some magnitude observations from the example data set,
which also includes the limiting magnitude.

```{r, echo=TRUE, results='hide'}
observations <- with(PER_2015_magn$observations, {
    idx <- !is.na(lim_magn) & sl_start > 135.81 & sl_end < 135.87
    data.frame(
        magn_id = magn_id[idx],
        lim_magn = lim_magn[idx]
    )
})
head(observations, 5) # Example values
```

```{r, echo=FALSE, results='asis'}
knitr::kable(head(observations, 5))
```

Next, the observed meteor magnitudes are matched with the corresponding observations.
This is necessary as we need the limiting magnitudes of the observations to determine
the r-value.

Using

```{r, echo=TRUE, results='hide'}
magnitudes <- with(new.env(), {
    magnitudes <- merge(
        observations,
        as.data.frame(PER_2015_magn$magnitudes),
        by = "magn_id"
    )
    magnitudes$magn <- as.integer(as.character(magnitudes$magn))
    subset(magnitudes, (magnitudes$lim_magn - magnitudes$magn) > -0.5)
})
head(magnitudes, 5) # Example values
```

we obtain a data frame with the absolute observed frequencies `Freq` for each
observation of a magnitude class. The expression
`subset(magnitudes, (magnitudes$lim_magn - magnitudes$magn) > -0.5`
ensures that meteors fainter than the limiting magnitude are not used if they exist.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(magnitudes[magnitudes$Freq > 0, ], 5))
```

This data frame contains a total of `r sum(magnitudes$Freq)` meteors.
This is a sufficiently large number to estimate the r-value.


## Maximum Likelihood Method

The maximum likelihood method can be used to estimate the r-value in an asymptotically
unbiased manner. For this, the function `dvmgeom()` is needed, which returns the
probability density of the observable meteor magnitudes when the r-value and the
limiting magnitudes are known.

The following algorithm estimates the r-value by maximizing the likelihood with
the `optim()` function. The function `ll()` returns the negative log-likelihood,
as `optim()` identifies a minimum.

```{r, echo=TRUE, results='hide'}
# maximum likelihood estimation (MLE) of r
result_ml <- with(magnitudes, {
    # log likelihood function
    ll <- function(r) -sum(Freq * dvmgeom(magn, lim_magn, r, log = TRUE))
    r_start <- 2.0 # starting value
    r_lower <- 1.2 # lowest expected value
    r_upper <- 4.0 # highest expected value
    # find minimum
    optim(r_start, ll, method = "Brent", lower = r_lower, upper = r_upper, hessian = TRUE)
})
```

This gives the expected value and the variance of the r-value:

```{r, echo=TRUE}
print(result_ml$par) # mean of r
print(1 / result_ml$hessian[1][1]) # variance of r
```

We can additionally visualize the likelihood function here.

```{r, echo=TRUE, results='hide'}
with(new.env(), {
    data_plot <- data.frame(r = seq(2.0, 2.8, 0.01))
    data_plot$ll <- mapply(function(r) {
        with(magnitudes, {
            # log likelihood function
            sum(Freq * dvmgeom(magn, lim_magn, r, log = TRUE))
        })
    }, data_plot$r)
    data_plot$l <- exp(data_plot$ll - max(data_plot$ll))
    data_plot$l <- data_plot$l / sum(data_plot$l)
    brks <- seq(min(data_plot$r) - 0.02, max(data_plot$r) + 0.02, by = 0.02)
    plot(data_plot$r, data_plot$l,
        # breaks = brks,
        type = "l",
        col = "blue",
        xlab = "r",
        xaxt = "n",
        ylab = "likelihood"
    )
    xlabels <- seq(min(round(data_plot$r, 1)) - 0.1, max(round(data_plot$r, 1)) + 0.1, by = 0.1)
    axis(
        side = 1,
        at = xlabels,
        labels = sprintf("%.1f", xlabels)
    )
    abline(v = result_ml$par, col = "red", lwd = 1)
})
```

The likelihood function is approximately a normal distribution in this case. This is important in this context because the variance of the estimated r-value is derived from the curvature (the second derivative at the maximum) of the log-likelihood function.


## Generalized Linear Model

The maximum likelihood estimate can also be obtained as a generalized linear
model. `vmgeom_glm()` fits the same likelihood through `glm()`, which becomes
useful as soon as the r-value is no longer assumed to be constant.

The response is passed as a two-column matrix holding the meteor magnitude and
the corresponding limiting magnitude. Supplying the limiting magnitude as part
of the response ensures that it is subsetted together with the remaining data
whenever `glm()` drops rows. The observed frequencies enter as `weights`.

```{r, echo=TRUE, results='hide'}
# rows with a frequency of zero carry no information
magnitudes_obs <- subset(magnitudes, magnitudes$Freq > 0)

result_glm <- vmgeom_glm(
    cbind(magn, lim_magn) ~ 1,
    data = magnitudes_obs,
    weights = magnitudes_obs$Freq
)
```

The link function is $\eta = \mathrm{logit}(1/r) = -\log(r-1)$. The quantity $1/r$ is
the factor by which the observable rate drops when the limiting magnitude is lowered
by one, and therefore lies between $0$ and $1$, which makes a logit link the natural
choice. It also enforces $r > 1$ for every finite $\eta$, a constraint the geometric
model requires.

Rather than back-transforming the coefficients by hand, `predict()` returns the
population index directly. Both methods agree, which is expected: without
covariates this is an intercept-only fit of the same likelihood.

```{r, echo=TRUE}
# the model has no covariates, so any single row predicts the global r-value
newdata <- data.frame(row.names = "")
print(c(
    "ML" = result_ml$par,
    "GLM" = as.numeric(predict(result_glm, newdata, type = "r"))
))
```

The same holds for the uncertainty. The maximum likelihood standard deviation
comes from the curvature of the log-likelihood, the one of the linear model from
the standard error of $\eta$, transformed with the delta method:

```{r, echo=TRUE}
print(c(
    "ML" = sqrt(1 / result_ml$hessian[1][1]),
    "GLM" = as.numeric(predict(result_glm, newdata, type = "r", se.fit = TRUE)$se.fit)
))
```

For a single r-value the likelihood approach is the cheaper one. On the data set
used here it is roughly twice as fast as the linear model, since `glm()` carries
the overhead of its iterative fitting on top of the same likelihood. Both are a
matter of milliseconds, and for aggregated frequencies the cost is governed by
the number of magnitude classes rather than by the number of meteors.

Predictions are available on several scales — `"r"`, `"inv_r"`, `"log_r"` and
`"link"` — and `se.fit = TRUE` adds confidence limits in addition to the
standard error. The limits are formed on the link scale and then transformed, so
they never fall below `1`.

```{r, echo=TRUE}
print(predict(result_glm, newdata, se.fit = TRUE))
```

The real benefit appears once covariates are added. Because the fit is an
ordinary `glm` object, `summary()`, `anova()`, `AIC()`/`BIC()` and
`select_knots()` can all be used. A model in which the r-value varies with
solar longitude would be written as

```r
vmgeom_glm(cbind(magn, lim_magn) ~ sl, data = magnitudes, weights = Freq)
```

and a smooth activity-dependent profile as

```r
vmgeom_glm(cbind(magn, lim_magn) ~ splines::ns(sl, df = 3), data = magnitudes, weights = Freq)
```

Since the family reports the exact log-likelihood of the fitted distribution, `AIC()` also
allows a comparison against a different magnitude distribution altogether. Fitting the ideal
distribution to the same data with `vmideal_glm()` gives:

```{r, echo=TRUE}
result_ideal <- vmideal_glm(
    cbind(magn, lim_magn) ~ 1,
    data = magnitudes_obs,
    weights = magnitudes_obs$Freq
)
print(c("geometric" = AIC(result_glm), "ideal" = AIC(result_ideal)))
```

Each model has one parameter, so the comparison reduces to their likelihoods. The ideal
distribution attains the lower value here, though by a margin that this excerpt of
`r sum(magnitudes$Freq)` meteors does not resolve — a difference of a few units is weak
evidence, and the two models are hard to tell apart at the limiting magnitudes involved.
Such a comparison is only valid when both fits use the same observations and the same
response.

Covariates can also be accommodated on a different basis, by transforming the
magnitudes first; see the section on the variance-stabilizing transformation
below.


## By Rate

This method asks how strongly the observed rate would drop if the limiting magnitude were
lowered by one. Under the model of the introduction, the expected number of meteors
recorded at limiting magnitude $m_{\mathrm{lim}}$ is proportional to

$$
S(m_{\mathrm{lim}}) \;=\; \sum_{m \in \mathbb{Z}} f(m_{\mathrm{lim}} - m)\, r^{m},
$$

each magnitude class contributing its population share $r^{m}$ weighted by the probability
$f(\cdot)$ of actually perceiving it. Lowering the limiting magnitude by one and forming
the ratio gives the relative drop of the rate,

$$
\frac{S(m_{\mathrm{lim}} - 1)}{S(m_{\mathrm{lim}})} \;=\; \frac{1}{r},
$$

which turns out to be exactly $1/r$ — independent of $m_{\mathrm{lim}}$ and of the shape of
$f(\cdot)$. Measuring this drop is what gives the method its name.

The rates themselves are not observed, but the same ratio can be formed per meteor. For a
meteor of magnitude $m$ observed at limiting magnitude $m_{\mathrm{lim}}$ let

$$
a(m, m_{\mathrm{lim}}) \;=\; \frac{f(m_{\mathrm{lim}} - m - 1)}{f(m_{\mathrm{lim}} - m)},
$$

the factor by which its chance of being perceived would shrink if the limiting magnitude
were lowered by one, and write $a_i = a(m_i, m_{\mathrm{lim},i})$ for the $i$-th observed
meteor. This leads to

$$
\mathbb{E}\!\left[\frac{1}{n}\sum_{i=1}^{n} a_i\right] = \frac{1}{r},
$$

so averaging $a$ over the observed meteors estimates $1/r$ without bias. The derivation
below proves both identities.

### Derivation

**Setting.** Fix a limiting magnitude $m_{\mathrm{lim}}$ and let $M$ denote the magnitude
of a randomly selected observed meteor. Its distribution is

$$
P[M = m] \;=\; \frac{f(m_{\mathrm{lim}} - m)\, r^{m}}{S(m_{\mathrm{lim}})}.
$$

Two properties of the perception probability are used, and nothing else:

1. $f(x) = 0$ for $x \le -0.5$, that is, a meteor fainter than the limiting magnitude is
   never perceived. Hence $f(m_{\mathrm{lim}} - m) = 0$ for all $m \ge m_{\mathrm{lim}} +
   0.5$, and the sum is bounded above.
2. $0 \le f \le 1$. Together with $r > 1$ this bounds the sum from below as well, since
   $\sum_{m \le 0} r^{m}$ converges. So $0 < S(m_{\mathrm{lim}}) < \infty$ and the
   distribution is proper.

Let $m_{\max}$ be the largest class with $f(m_{\mathrm{lim}} - m_{\max}) > 0$. The support
of $M$ is contained in $\{m \le m_{\max}\}$, and on it the denominator of
$a(m, m_{\mathrm{lim}})$ is strictly positive, so $a(M, m_{\mathrm{lim}})$ is well defined
almost surely.

**Claim.** $\mathbb{E}\bigl[a(M, m_{\mathrm{lim}})\bigr] = 1/r$ for every
$m_{\mathrm{lim}}$.

**Proof.** Writing out the expectation and cancelling the perception probability, which is
legitimate because it is non-zero on the support,

$$
\begin{aligned}
\mathbb{E}\bigl[a(M, m_{\mathrm{lim}})\bigr]
 &= \sum_{m \le m_{\max}} \frac{f(m_{\mathrm{lim}} - m)\, r^{m}}{S(m_{\mathrm{lim}})}
    \cdot \frac{f(m_{\mathrm{lim}} - m - 1)}{f(m_{\mathrm{lim}} - m)} \\[5pt]
 &= \frac{1}{S(m_{\mathrm{lim}})} \sum_{m \le m_{\max}} f(m_{\mathrm{lim}} - m - 1)\, r^{m}.
\end{aligned}
$$

The remaining sum is by definition $S(m_{\mathrm{lim}} - 1)$, so that
$\mathbb{E}\bigl[a(M, m_{\mathrm{lim}})\bigr] = S(m_{\mathrm{lim}} - 1) / S(m_{\mathrm{lim}})$: the expectation of $a$
is precisely the relative drop of the rate stated above.

Before shifting the index, note that its uppermost term already vanishes. At $m = m_{\max}$
the numerator is evaluated one class beyond the faintest perceptible one,

$$
f\bigl(m_{\mathrm{lim}} - m_{\max} - 1\bigr) = 0,
$$

so the summation effectively stops at $m_{\max} - 1$. This is the step where the vanishing
perception probability enters, and it is what makes the shift below land exactly on the
range of $S(m_{\mathrm{lim}})$. Replacing $m$ by $m - 1$ therefore gives

$$
\begin{aligned}
\sum_{m \,\le\, m_{\max}} f(m_{\mathrm{lim}} - m - 1)\, r^{m}
 &= \sum_{m \,\le\, m_{\max} - 1} f(m_{\mathrm{lim}} - m - 1)\, r^{m} \\[5pt]
 &= \sum_{m \,\le\, m_{\max}} f(m_{\mathrm{lim}} - m)\, r^{m - 1} \\[5pt]
 &= \frac{1}{r} \sum_{m \,\le\, m_{\max}} f(m_{\mathrm{lim}} - m)\, r^{m} \\[5pt]
 &= \frac{S(m_{\mathrm{lim}})}{r},
\end{aligned}
$$

where the last sum is $S(m_{\mathrm{lim}})$ because $f(m_{\mathrm{lim}} - m) = 0$ for every
$m > m_{\max}$, so extending it to all integers adds nothing. Both series converge
absolutely, since $r^{m} \to 0$ as $m \to -\infty$, which is what permits the shift.
Altogether

$$
\begin{aligned}
\mathbb{E}\bigl[a(M, m_{\mathrm{lim}})\bigr]
 &= \frac{1}{S(m_{\mathrm{lim}})} \cdot \frac{S(m_{\mathrm{lim}})}{r} \\[5pt]
 &= \frac{1}{r}. \qquad \blacksquare
\end{aligned}
$$

Along the way this also proves the rate identity claimed above, since the intermediate
result was $S(m_{\mathrm{lim}} - 1) = S(m_{\mathrm{lim}})/r$.

Finally, the second identity follows. For $n$ meteors observed under limiting magnitudes
$m_{\mathrm{lim},1}, \dots, m_{\mathrm{lim},n}$, each $a_i$ satisfies
$\mathbb{E}[a_i] = 1/r$ by the claim, since it holds for *every* limiting magnitude
separately. Linearity of the expectation then gives

$$
\mathbb{E}\!\left[\frac{1}{n}\sum_{i=1}^{n} a_i\right] = \frac{1}{r}.
$$

### Properties of the estimator

The sample mean of $a$ is thus an unbiased estimator of $1/r$ — exactly, not merely
asymptotically, and irrespective of whether the observations share the same limiting
magnitude. This is what allows observations made under different conditions to be pooled.
The estimator is inexpensive to compute and directly comparable to the likelihood-based
approaches discussed earlier.

Unbiasedness for $1/r$ does not carry over to $r$ itself, however: $1/\bar{a}$ is a
non-linear function of $\bar{a}$, which is why the delta method is applied below.

A second property concerns the bright end. Once a meteor is bright enough that both
perception probabilities are close to `1.0`, the ratio approaches `1` as well, and it no
longer matters *how* bright the meteor actually was. At a limiting magnitude of `6.0` the
ratio is already `0.999` for a magnitude $-3$ meteor and indistinguishable from `1` for
anything brighter. Such meteors are therefore effectively pooled into a single class, and
the information about their exact brightness is lost.

In practice this is not a drawback but an advantage. Meteors that bright are rare — at
$r = 2.5$ and the same limiting magnitude, magnitude $-3$ and brighter accounts for less
than one percent of all observed meteors — so little information is discarded. In exchange, the pooling keeps a
single very bright meteor from acting as an outlier, precisely in the part of the
magnitude range where one observation would otherwise carry a lot of weight. The same
applies to the variance-stabilizing transformation below, which is built on this ratio.

### Estimation

We estimate $r$ as $1/\bar{a}$ from the sample mean $\bar{a}$. Since $r$ is a non-linear
function of $\bar{a}$, we apply the delta method to correct for bias and to compute the
variance of $r$. The delta method uses a Taylor expansion around the mean of $a$ to
approximate the distribution of $r$. Note that the variance of $a$ entering it depends on
$r$ itself; the section on the variance-stabilizing transformation below starts from this
same statistic and removes that dependency.

```{r, echo=TRUE, results='hide'}
result_rate <- with(magnitudes, {
    N <- sum(Freq)
    a <- vmperception(lim_magn - magn - 1) / vmperception(lim_magn - magn)
    a_mean <- as.numeric(weighted.mean(a, w = Freq))
    a_var <- as.numeric(cov.wt(cbind(a), wt = Freq)$cov) / N
    # apply the delta method and return the result
    list(
        "mean" = 1 / a_mean - a_var / a_mean^3,
        "var" = a_var / a_mean^4
    )
})
```

This gives the expected value and the variance of the r-value:

```{r, echo=TRUE}
print(result_rate$mean) # mean of r
print(result_rate$var) # variance of r
```

Using the bootstrap method, it can be assessed whether the mean is normally distributed.

```{r, echo=TRUE, results='hide'}
# Bootstrapping Method
r_means <- with(magnitudes, {
    N <- sum(Freq)
    a <- vmperception(lim_magn - magn - 1) / vmperception(lim_magn - magn)
    replicate(50000, {
        s <- sample(a, size = N, replace = TRUE, prob = Freq)
        s_mean <- mean(s)
        s_var <- var(s) / N
        1 / s_mean - s_var / s_mean^3
    })
})
```

The graphical representation indicates that this is indeed approximately the case.

```{r, echo=TRUE, results='show'}
with(new.env(), {
    r_sd <- sqrt(result_rate$var)
    r_min <- result_rate$mean - 3 * r_sd
    r_max <- result_rate$mean + 3 * r_sd
    r <- subset(r_means, r_means > r_min & r_means < r_max)
    brks <- seq(min(r) - 0.02, max(r) + 0.02, by = 0.02)
    hist(r,
        breaks = brks,
        col = "skyblue",
        border = "black",
        main = "Histogram of mean r",
        xlab = "r",
        xaxt = "n",
        ylab = "count"
    )
    xlabels <- seq(min(round(r, 1)) - 0.1, max(round(r, 1)) + 0.1, by = 0.1)
    axis(
        side = 1,
        at = xlabels,
        labels = sprintf("%.1f", xlabels)
    )
    abline(v = result_rate$mean, col = "red", lwd = 1)
})
```


## Variance-Stabilizing Transformation as an Alternative Method

This method is a refinement of the rate-based one. It uses the same quantity
$a(m, m_{\mathrm{lim}})$, but rescales it so that its variance no longer depends on the
parameter `r`. It is likewise far cheaper than estimation based on the maximum likelihood
principle.

Recall what was established in the previous section. The sample mean of $a$ estimates
$1/r$ without bias, exactly and for pooled limiting magnitudes alike, which is a strong
property. Its weakness is the second moment: the variance of $a$ is itself a function of
$r$. At a limiting magnitude of `6.0` it falls from about `0.066` at $r = 1.4$ to `0.033`
at $r = 4$, a factor of two across the range of interest. Two consequences follow. The
precision of the estimate cannot be judged without already knowing $r$, and the delta
method is needed to convert $\bar{a}$ into an estimate of $r$ at all.

The variance-stabilizing transformation removes exactly this weakness. It maps $a$ onto a
scale on which the variance is approximately `1.0` regardless of $r$ — over
$1.4 \le r \le 4$ it stays within about `0.16` of that value, comfortably covering the
$1.7 \le r \le 3.3$ met in visual meteor work. Outside that window the stabilization
degrades on both sides, which bounds the use of the transformed magnitudes as a response
in linear models, but not the reading of a single $r$ off their mean.
`vmgeom_vst_from_magn()` performs the mapping and `vmgeom_vst_to_r()` maps the mean of the
transformed magnitudes back onto $r$.

The mapping is monotonically decreasing: a bright meteor is far from the limiting magnitude
and receives a large `tm`. Towards the bright end it runs into a ceiling, since $a$ cannot
exceed `1` and the transformed value is therefore bounded by what the transformation makes
of that limit. At the faint end it reaches zero half a magnitude below the limiting
magnitude, where a meteor is no longer perceived. Shown here for a limiting magnitude of
`6.0`.

```{r, echo=TRUE, results='show'}
with(new.env(), {
    lim_magn <- 6.0
    m <- seq(-2, 5.5, 0.05)
    plot(m, vmgeom_vst_from_magn(m, lim_magn),
        type = "l",
        col = "blue",
        xlab = "m",
        ylab = "tm"
    )
})
```

The way back needs no limiting magnitude at all, since the statistic it rests on does not
depend on one. On logarithmic axes the relation is nearly a straight line, which is the additivity
discussed further below.

```{r, echo=TRUE, results='show'}
with(new.env(), {
    tm <- seq(1.5, 4.3, 0.01)
    plot(tm, vmgeom_vst_to_r(tm),
        type = "l",
        col = "blue",
        log = "xy",
        xlab = "tm",
        ylab = "r",
        xaxt = "n",
        yaxt = "n"
    )
    axis(1, at = c(1.5, 2, 3, 4), labels = c("1.5", "2", "3", "4"))
    axis(2, at = c(1, 1.5, 2, 3, 4), labels = c("1", "1.5", "2", "3", "4"))
})
```

The price is paid on the first moment. The transformation is non-linear in $a$, and the
mean of a non-linear function is not that function of the mean, so the exact unbiasedness
of the rate-based estimator is lost. This deviation is systematic: it does not average out,
and it does not shrink as the sample grows.

Its cause is a slight dependence of the transformed mean on the limiting magnitude, which
the statistic $a$ itself does not have. Over $1.7 \le r \le 3.3$, the range met in visual
meteor work, it amounts to at most `0.8%` of $r$.

What matters in practice is how that compares with the random error, and this depends on
the number of meteors. The random error shrinks as the sample grows while the systematic
one does not, so the two meet at a certain sample size: for a few hundred meteors the
random error still dominates by a factor of five or more, at a few thousand by roughly
two, and only beyond some ten thousand does the systematic part set the accuracy — no
matter how many more meteors are added. Visual meteor data sets of that size are the
exception, so for most analyses the deviation can be ignored.

Where it does matter — for a very large data set, or wherever an unbiased estimate counts
for more than a constant variance — the rate-based method of the previous section remains
exactly unbiased for $1/r$, and `vmgeom_glm()` fits the exact likelihood.

Everything the rate-based method offers is retained. Since the transformation is a
monotone rescaling of $a$, the pooling of bright meteors described above carries over
unchanged. And because $a$ lies between `0` and `1`, the transformed values are bounded as
well, with the upper bound corresponding to $r = 1$; the estimate can consequently never
fall below the value the geometric model requires.

The same monotonicity matters when the transformation is used inside a predictive model.
`vmgeom_vst_to_r()` is defined on the whole range of values the transformation can produce
and returns `NA` only for values it cannot. Should sparse data drive a prediction into an
implausible region, the result is therefore a value that is too large but still ordered,
rather than a missing one — usually far easier to handle downstream than an `NA`.

The delta method is applied below in this section as well, since the back-transformation
is non-linear here too. What changes is the variance that enters it: a constant one
instead of a value that depends on the very parameter being estimated.

Beyond stabilizing the variance, the transformation also puts the parameter on a scale
with a simple structure. Classically, transformations of this kind are judged by three
criteria, which Box and Cox (1964, *An Analysis of Transformations*, J. R. Stat. Soc. B
26, 211–252) treat together: constant variance, an approximately symmetric distribution,
and *additivity*, meaning that effects combine linearly on the transformed scale. The
third one is what the transformation is built around. It belongs to the Box-Cox family
itself, applied to $a$, and $a$ estimates $1/r$, so

$$
\log t \;=\; \mathrm{const.} - \beta \, \log r
$$

with $\beta > 0$: the relation is close to linear on the logarithmic scale. A multiplicative
change of the population index therefore becomes an approximately additive shift of
$\log t$. Over the range of $r$ the calibration covers this is what the mean of the
transformed magnitudes follows, to the accuracy stated above, and it is what the
calibration of `vmgeom_vst_to_r()` rests on. The departure from exact linearity is slight
but systematic, and the back-transformation accounts for it explicitly — see below.

This is what makes the transformed magnitudes usable as a response in ordinary linear
models, with residuals that are homoscedastic by construction — precisely what such models
assume. `vmgeom_vst_lm()` fits them, and the estimation below is already of that form:
taking the mean of the transformed magnitudes *is* an intercept-only linear model. A
profile in which $r$ varies with a covariate only extends the right-hand side of the
formula, so that

```r
vmgeom_vst_lm(cbind(magn, lim_magn) ~ sl, data = magnitudes, weights = Freq)
```

fits a solar-longitude dependence in the same way `vmgeom_glm()` does above. The estimated
coefficients translate back into changes of $\log r$ up to the factor $\beta$. The
back-transformation is $\log r = c + d \log t + e\,t$, so that factor is
$\beta = -1/(d + e\,t)$, which over the range met in practice runs from about `0.73` to
`0.81`. The term in $t$ is what captures the curvature of the relation; a straight line
through it would leave a systematic error about `1.7` times as large, and it is the reason
`vmgeom_vst_to_r()` reproduces $r = 1$ exactly at the upper bound of `tm`.

That makes this approach a lightweight alternative to `vmgeom_glm()`, and the choice
between them is a trade-off. In favour of the transformation is its cost: on the data set
used here a single estimate takes well under a millisecond, some two orders of magnitude
less than the generalized linear model, which iterates over the full likelihood. Since the
transformation is applied once per meteor and everything afterwards is ordinary least
squares, this advantage grows with the size of the data set, and the whole toolbox built
around `lm()` becomes available.

Against it stands a loss of accuracy, though not where one would expect it. The dispersion
of the estimate is practically that of the maximum likelihood estimate — in a simulation
with `n = 500` the two standard deviations agree to within a few percent over
$2 \le r \le 4$. What the transformation costs is not variance but the systematic
deviation of the back-transformation described above, which adds to the total error and,
unlike the random part, does not shrink with the sample size. At `n = 500` and $r = 3$ it
raises the root mean squared error by roughly `3%` over that of the maximum likelihood
estimate; for larger samples its share grows. `vmgeom_glm()`, fitting
the exact likelihood, carries no such deviation and additionally provides the
inferential machinery of a `glm` object — `summary()`, `anova()`, information criteria.
The transformation is therefore preferable for quick estimates and exploratory work, the
generalized linear model whenever accuracy matters — including, unlike the usual
trade-off, for large data sets, where the systematic part comes to dominate.

A constant variance carries a few further consequences worth noting:

- Standard errors and confidence intervals no longer depend on the unknown r-value, so they are equally well calibrated across the entire parameter space,
- Since the sample mean is homoscedastic, the classical results of linear estimation theory (BLUE properties, Gauss–Markov theorem) apply without further transformation,
- Test statistics such as likelihood ratio or score tests follow a more uniform distribution, because the variance need not be treated as an additional unknown. This improves test power and simplifies asymptotic approximations.

The resulting procedure is straightforward: all that is needed is the mean of the
transformed meteor magnitudes, from which an estimate of the parameter `r` is obtained.
`vmgeom_vst_lm()` does both in one step. It applies the transformation and fits the result
as the intercept-only linear model announced above, so that the step to a covariate is a
matter of extending the formula. The response is passed as `cbind(magn, lim_magn)` and the
observed frequencies as `weights`, exactly as for `vmgeom_glm()`.

```{r, echo=TRUE, results='show'}
result_vs <- vmgeom_vst_lm(
    cbind(magn, lim_magn) ~ 1,
    data = magnitudes,
    weights = magnitudes$Freq
)
```

`predict()` converts the fitted mean back to the `r` scale. It applies the delta method,
which accounts for the nonlinearity of the back-transformation, and returns the standard
error alongside the estimate.

```{r, echo=TRUE, results='show'}
# the model has no covariates, so any single row predicts the global r-value
newdata <- data.frame(row.names = "")
prediction <- predict(result_vs, newdata, se.fit = TRUE, bias_correction = TRUE)
```

Two of the arguments deserve a comment. Passing the frequencies as `weights` is the
correct and complete way to supply aggregated counts — nothing is lost by it, and there is
no separate mechanism one ought to be using instead. The coefficients it produces are
exactly those of a data set holding one row per meteor.

What differs is only how the residual variance is *read back*. R has no notion of
frequency weights in `lm()`: weights are taken to express precision, so the residual
variance is normalized by the number of rows. `summary()` therefore reports a standard
error that refers to the `r nrow(magnitudes[magnitudes$Freq > 0, ])` magnitude classes
rather than to the `r sum(magnitudes$Freq)` meteors behind them — here about `40%` too
large, with the confidence interval correspondingly too wide. The same applies to
`glm(family = gaussian())`.

`predict()` takes care of this: it derives the residual scale from the sum of the weights
and thus reports the standard error of the meteors. Nothing needs to be done differently
when calling the function — but the standard errors of `summary(result_vs)` should not be
used, since they carry the row-based normalization.

Expanding the rows into one meteor each would settle the question at the source, and for
integer counts it gives exactly the same result as the correction applied here. Visual
observations, however, may contain half meteors, which makes that route unavailable.

`bias_correction = TRUE` adds the second-order term of the delta method. It is not the
default — `predict()` methods in R report a first-order standard error, and so does
`vmgeom_glm()` above — but it is what this section is about. The term is small here, well
below the random error at this sample size, and it corrects the *curvature* of the
back-transformation, not the systematic deviation described further above. That deviation
is a property of the calibration and does not shrink as the sample grows.

Thus, one obtains the mean and the variance of the mean of `r`.

```{r, echo=TRUE, results='show'}
result_vs <- list(mean = unname(prediction$fit), var = unname(prediction$se.fit)^2)
print(paste("r mean:", result_vs$mean))
print(paste("r var:", result_vs$var))
```

Using the bootstrap method, it can be assessed whether the mean is normally distributed.

```{r, echo=TRUE, results='hide'}
# Bootstrapping Method
r_means <- with(magnitudes, {
    N <- sum(Freq)
    tm <- vmgeom_vst_from_magn(magn, lim_magn)
    replicate(50000, {
        s <- sample(tm, size = N, replace = TRUE, prob = Freq)
        s_mean <- mean(s)
        s_var <- var(s) / N

        r_hat <- vmgeom_vst_to_r(s_mean)
        d2r_ds2 <- vmgeom_vst_to_r(s_mean, deriv_degree = 2L)
        r_hat - 0.5 * d2r_ds2 * s_var
    })
})
```

The graphical representation indicates that this is indeed approximately the case.

```{r, echo=TRUE, results='show'}
with(new.env(), {
    r_sd <- sqrt(result_vs$var)
    r_min <- as.vector(result_vs$mean - 3 * r_sd)
    r_max <- as.vector(result_vs$mean + 3 * r_sd)
    r <- subset(r_means, r_means > r_min & r_means < r_max)
    brks <- seq(min(r) - 0.02, max(r) + 0.02, by = 0.02)
    hist(r,
        breaks = brks,
        col = "skyblue",
        border = "black",
        main = "Histogram of mean r",
        xlab = "r",
        xaxt = "n",
        ylab = "count"
    )
    xlabels <- seq(min(round(r, 1)) - 0.1, max(round(r, 1)) + 0.1, by = 0.1)
    axis(
        side = 1,
        at = xlabels,
        labels = sprintf("%.1f", xlabels)
    )
    abline(v = result_vs$mean, col = "red", lwd = 1)
})
```

## Residual Analysis

So far, we have operated under the assumption that the real distribution of meteor magnitudes
is exponential and that the perception probabilities are accurate.
We now use the Chi-Square goodness-of-fit test to check whether the observed frequencies match
the expected frequencies. Then, using the estimated r-value, we retrieve the relative
frequencies `p` for each observation and add them to the data frame `magnitudes`:

```{r, echo=TRUE, results='hide'}
magnitudes$p <- with(magnitudes, dvmgeom(m = magn, lm = lim_magn, result_rate$mean))
```

We must also consider the probabilities for the magnitude class with the brightest meteors.

```{r, echo=TRUE, results='hide'}
magn_min <- min(magnitudes$magn)
```

The smallest magnitude class `magn_min` is `r magn_min`. In calculating the probabilities,
we assume that the magnitude class `r magn_min` contains meteors that are either brighter
or equally bright as `r magn_min` and thus use the function `pvmgeom()` to determine
their probability.

```{r, echo=TRUE, results='asis'}
idx <- magnitudes$magn == magn_min
magnitudes$p[idx] <- with(
    magnitudes[idx, ],
    pvmgeom(m = magn + 1L, lm = lim_magn, result_rate$mean, lower.tail = TRUE)
)
```

This ensures that the probability of observing a meteor of any given magnitude is 100%.
This is known as the normalization condition. Accordingly, the Chi-Square goodness-of-fit test
will fail if this condition is not met.

We now create the contingency table `magnitutes_observed` for the observed meteor magnitudes
and its margin table.

```{r, echo=TRUE}
magnitutes_observed <- xtabs(Freq ~ magn_id + magn, data = magnitudes)
magnitutes_observed_mt <- margin.table(magnitutes_observed, margin = 2)
print(magnitutes_observed_mt)
```

Next, we check which magnitude classes need to be aggregated so that each contains
at least 10 meteors, allowing us to perform a Chi-Square goodness-of-fit test.

The last output shows that meteors of magnitude class `0` or brighter must be combined into
a magnitude class `0-`. Meteors with a brightness less than `4` are grouped here in the
magnitude class `4+`, and a new contingency table magnitudes.observed is created:

```{r, echo=TRUE}
magnitudes$magn[magnitudes$magn <= 0] <- "0-"
magnitudes$magn[magnitudes$magn >= 4] <- "4+"
magnitutes_observed <- xtabs(Freq ~ magn_id + magn, data = magnitudes)
print(margin.table(magnitutes_observed, margin = 2))
```

We now need the corresponding expected relative frequencies

```{r, echo=TRUE}
magnitutes_expected <- xtabs(p ~ magn_id + magn, data = magnitudes)
magnitutes_row_freq <- margin.table(magnitutes_observed, margin = 1)
magnitutes_expected <- sweep(magnitutes_expected, 1, magnitutes_row_freq, `*`)
magnitutes_expected <- magnitutes_expected / sum(magnitutes_expected)
print(sum(magnitudes$Freq) * margin.table(magnitutes_expected, margin = 2))
```

and then carry out the Chi-Square goodness-of-fit test:

```{r, echo=TRUE, results='asis'}
chisq_test_result <- chisq.test(
    x = margin.table(magnitutes_observed, margin = 2),
    p = margin.table(magnitutes_expected, margin = 2)
)
```

As a result, we obtain the p-value:

```{r, echo=TRUE}
chi2_df <- chisq_test_result$parameter - 1
chi2_pval <- pchisq(chisq_test_result$statistic, df = chi2_df, lower.tail = FALSE)
print(chi2_pval)
```

If we set the level of significance at 5 percent, then it is clear that the p-value with
`r unname(chi2_pval)` is greater than 0.05. Thus, under the assumption that the
magnitude distribution follows an geometric meteor magnitude distribution and assuming that
the perception probabilities are correct (i.e., error-free or precisely known),
the assumptions cannot be rejected. However, the converse is not true; the assumptions
may not necessarily be correct. The total count of meteors here is too small for such
a conclusion.

To verify the p-value, we also graphically represent the Pearson residuals:

```{r, fig.show='hold'}
chisq_test_residuals <- with(new.env(), {
    chisq_test_residuals <- residuals(chisq_test_result)
    v <- as.vector(chisq_test_residuals)
    names(v) <- names(chisq_test_residuals)
    v
})
plot(
    chisq_test_residuals,
    main = "Residuals of the chi-square goodness-of-fit test",
    xlab = "m",
    ylab = "Residuals",
    ylim = c(-3, 3),
    xaxt = "n"
)
abline(h = 0.0, lwd = 2)
axis(1, at = seq_along(chisq_test_residuals), labels = names(chisq_test_residuals))
```
